Efficient evaluation of threshold queries of derived fields in a numerical simulation database

### 2014-09-03 paper.tex
Added PDF of vorticity. Some reorganization.

\begin{abstract}
Large numerical simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and
even PB ranges. Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
interested in out of the box. In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets
stored in a cluster of relational databases. The method achieves scalability through data-parallel execution. We introduce an application-aware 
cache for query results, which has little overhead and improves query performance by X times for queries that hit the cache.

@@

\section{Introduction}
%What is the problem?
Better instruments, faster and bigger supercomputers and easier collaboration and sharing of data in the sciences have introduced the need
to manage increasingly large data sets. Data-intensive systems and architectures have been developed with the goal of storing and providing
fast access to such data sets. Examples include the GrayWulf and DataScope clusters \cite{Szalay, TODO} at Johns Hopkins, which have capacity 
of 1.1PB and xPB
respectively. These systems differ from the traditional HPC systems in that they aim to achieve high aggregate throughput by balancing computation
capabilities with I/O and network bandwidth. The computing systems and services developed on top of these platforms are more than pure
storage engines and usually have complex analysis routines built-in, which has largely been driven by the ``move the computation to the data"
paradigm. These built-in analysis routines are most often not novel themselves. They implement core scientific functionality for the study of the
particular scientific phenomena, which was observed or simulated in the first place. The analysis routines however require novel evaluation strategies 
and methods for their execution. They have to operate on large datasets distributed across multiple nodes of a cluster of relational databases 
and in order to reduce their running times they have to make efficient use of the cluster resources.

@@

%Why is it hard? (E.g., why do naive approaches fail?)
The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
In order to evaluate such a query for a given time-step all of the data for that time-step must be read and each data point 
will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.
It is impractical to materialize all possible derived fields due to the large size of the datasets and the limits of available storage. Therefore, the derived fields 
and threshold queries on them have to be evaluated on-demand from the stored data.
Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, as was
shown by Lopez et al. \cite{Lopez} this might not be sufficient for some interactive applications. Moreover, even if the data are available in the cache the
computation associated with the derived field still needs to be performed for each point on the grid as results of previous computations are not cached.

@@

%What are the key components of my approach and results? Also include any specific limitations.
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations. Each computation is distributed across the nodes of the cluster and
is executed in parallel by several threads within a database node by means of creating multiple connections to each node. Moreover we introduce an 
application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
on the concept of semantic caching \cite{Ren}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.
we cache the actual results of the computation, which leads to query performance improvement of up to X times.

@@

\section{Scientific Use Cases}\label{science_use_cases}
Max vorticity clustering.

@@

{\bf Distributed data-parallel execution}. In most cases threshold queries operate over an entire time-step. Therefore, each such query is subdivided by 
the mediator into queries submitted to
each of the database nodes. Each node evaluates the query over the data that it has stored locally. Only a small amount of data along the boundary
need to be requested from adjacent nodes. The data are read into memory and the particular field requested is computed at each of the locations on
the grid. The same strategy applies when utilizing multiple threads per node. The norm or absolute value of the derived field at each location is compared 
against the specified threshold and if the value is higher it is maintained along with
the spatial coordinates of the location in a list. We limit the maximum size of this list to around 1,000,000 locations in order to prevent having to return
the entire time-step or a significant fraction of it for queries with thresholds that are set too low.

{\bf Application-aware cache for query results}. The central part of the evaluation strategy for threshold queries that we have developed is the 
introduction of an application-aware cache
for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Each database node has a local 
cache. Cache entries are indexed by the field, time-step, spatial region and the threshold requested. We use a least recently
used cache replacement policy. All modifications of and queries to the cache are executed within a transaction with snapshot isolation level to
avoid dirty-reads or inconstant view of the cache. Caching the query results
preserves the computational effort in addition to substantially reducing I/O. We do not have to derive the requested field from the raw data for 
queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
a small set of data and do not need to perform any additional computation.

@@

\begin{table}
\centering
\begin{tabular}{|c|c|c|c|} \hline
\multirow{2}{0.8in}{Points above threshold} & \multicolumn{3}{|c|}{Running time (s.)}\\ \cline{2-4}
 & No cache & \parbox[t]{0.7in}{With cache (miss)} & \parbox[t]{0.7in}{With cache (hit)} \\ \hline
4247 & 97.1 & 100.2 & 0.5\\ \hline
86580 & 113.7 & 115.9 & 1.2\\ \hline
909274 & 111.6 & 115.0 & 9.1\\ \hline
\end{tabular}
\caption{Effectiveness of caching.}
\label{tab:cache_effectiveness}
\end{table}

The experiments were run on the production database nodes of the JHTDB through a development Web-server hosting the Web-services. We used
the MHD dataset for the experimental runs. This dataset is partitioned across 4 database nodes according to spatial regions in the Morton z-order. 
The database nodes are 2.66 GHz dual quad-core Windows 2008 servers with SQL Server 2008 R2 and 24 GB of memory. Each node has 24 2TB 
SATA disks arranged as a set of four RAID-5 arrays. The database files are distributed in a round-robin fashion across the nodes and their associated disk
arrays. The tables storing the data are partitioned spatially along contiguous ranges of the Morton z-curve and the data for each partition reside in one
database file.

For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
have to be computed from the stored data. The vorticity is computed from the velocity field by taking its curl:
\begin{align}
\vec{\omega} = \vec{\nabla} \times \vec{v}.
\label{eq:curl}
\end{align}
We use finite differencing methods of
different orders for the evaluation of the curl. As described in section \ref{science_use_cases}
thresholding the vorticity field is important in the study of fluid dynamics and clustering the locations of maximum vorticity can lead to new insights into
the development of the most intense vortices that are observed. \kk{Consider showing other fields as well; possibly thresholding the velocity by itself}

\begin{figure}
\centering
\includegraphics[width=3.5in]{Figures/cache_overhead.pdf}
\caption{Execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the absence of a cache.}
\label{fig:cache_overhead}
\end{figure}

The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high and only $\sim$4,300 points 
(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold was chosen and $\sim$87,000 points 
(or 0.0081\% of all points) were above the threshold. Finally, a low threshold was chosen for the last set with $\sim$900,000 points 
(or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user. 
The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
Cache entries for the particular time-step queried were dropped before each run, making sure that each query would produce a cache miss and would
have to be evaluated from the raw data. 
The execution times were averaged over 10 runs.
We utilized 4 threads per database
node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is minimal, 
less than 3\% and within
the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels. The slight increase in the running
time that can be observed is due to the larger result set that has to be returned to the user.
\kk{Consider including a figure for a representative time-step showing the PDF of the vorticity field
to show how the vorticity values are distributed and to get an understanding for the different threshold levels.}

\begin{figure}
\centering
\includegraphics[width=3.5in]{Figures/scaleup.pdf}
\caption{Execution time for threshold queries at different threshold levels -- high, medium and low. The evaluation was performed utilizing 1-8
threads per server showcasing the vertical scalability of the implementation.}
\label{fig:scaleup}
\end{figure}

@@

The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we used the same
queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} but with varying number of threads per node. Cache entries for the
time-step queried were again dropped before each run in order to evaluate the scaling properties of the computation of the derived field from the stored data.
The computations for all of 
the derived fields of interest (such as the vorticity) at each gird point need data from adjacent grid points only. Therefore, each node of the cluster is able to
compute the derived field from data available locally with only a little amount along the boundary of each region having to be retrieved from adjacent nodes.
Each computation is independent and can be executed in parallel. This allows us to make use of multiple threads per node.

@@

\begin{figure}
\centering
\includegraphics[width=3.5in]{Figures/io_comparison.pdf}
\caption{Execution time for threshold queries evaluated utilizing different number of
threads per server compared with the time taken to perform the I/O only.}
\label{fig:io_comparison}
\end{figure}

>>>>

\begin{abstract}
Large numerical simulations are commonplace in the study of many different physical phenomena. 
The datasets produced by these simulations are in the TB and
even PB ranges. Data-intensive architectures and compute clusters have been deployed to store, manage and provide public access to such datasets. 
The systems built on top of these architectures, sometimes called virtual laboratories, usually do not support the analysis functionality that scientists are 
interested in out of the box. Moreover, data-intensive computations that examine entire time-steps of the simulation data are impractical to perform
locally and can take days or months to iterate over the entire dataset. 
In this paper we present a method for the efficient evaluation of threshold queries of derived fields for large numerical 
simulation datasets
stored in a cluster of relational databases. The method achieves scalability through data-parallel execution. We introduce an application-aware 
cache for query results, which has little overhead and improves query performance by over an order of magnitude for queries that hit the cache.

@@

\section{Introduction}
%What is the problem?
Better instruments, faster and bigger supercomputers and easier collaboration and sharing of data in the sciences have introduced the need
to manage increasingly large data sets. Data-intensive systems and architectures have been developed with the goal of storing and providing
fast access to such data sets. Examples include the GrayWulf and DataScope clusters \cite{Szalay, TODO} at Johns Hopkins, which have capacity 
of 1.1PB and xPB
respectively. They have been developed with the goal of providing persistent storage and public access to world-class numerical simulation data. 
These systems differ from the traditional HPC systems in that they aim to achieve high aggregate throughput by balancing computation
capabilities with I/O and network bandwidth. The computing systems and services developed on top of these platforms are more than pure
storage engines and usually have complex analysis routines built-in, which has largely been driven by the ``move the computation to the data"
paradigm. These built-in analysis routines are most often not novel themselves. They implement core scientific functionality for the study of the
particular scientific phenomena, which was observed or simulated in the first place. The analysis routines however require novel evaluation strategies 
and methods for their execution. They have to operate on large datasets distributed across multiple nodes of a cluster of relational databases 
and in order to reduce their running times they have to make efficient use of the cluster resources.

@@

%Why is it hard? (E.g., why do naive approaches fail?)
The evaluation of threshold queries of derived fields requires a kernel computation at each point on the grid in order to compute the field to be thresholded.
The value of the derived field at each location depends on all the values at locations within the kernel of computation. 
In order to evaluate such a query for a given time-step all of the data for that time-step must be read and each data point 
will appear in $N$ kernels of computation, where $N$ is the size of the kernel. The evaluation is thus both data and computation intensive.
Obtaining the derived field and thresholding locally requires not only the computation of the derived field over an entire time-step but also the transfer of
a large amount of data most of which are subsequently discarded. 
It is also impractical to materialize all possible derived fields with the raw data
due to the large size of the datasets and the limits of available storage. Therefore, the derived fields 
have to be evaluated on-demand from the stored data and thresholding performed on the database nodes near the data.

Database, operating and file system caches are effective at speeding up access to the large amounts of data stored on disk. However, as was
shown by Lopez et al. \cite{Lopez} this might not be sufficient for some interactive applications. Moreover, even if the data are available in the cache the
computation associated with the derived field still needs to be performed for each point on the grid as results of previous computations are not cached.

@@

%What are the key components of my approach and results? Also include any specific limitations.
We exploit the parallelism 
available in the cluster by means of data-parallel execution of the computations. Each computation is distributed across the nodes of the cluster and
is executed in parallel by several threads within a database node by means of creating multiple connections to each node. Moreover we introduce an 
application-aware cache for query results. We build on the idea of an application-aware cache introduced by Lopez et al. \cite{Lopez} and more broadly 
on the concept of semantic caching \cite{Ren}. Rather than caching just data as is the case in system caches and the tree-cache described by Lopez et al.
we cache the actual results of the computation, which leads to query performance improvement of over an order of magnitude.

@@

\section{Scientific Use Cases}\label{science_use_cases}
%Max vorticity clustering.
One of the applications of thresholding in turbulence is to find the locations of maximum vorticity in a particular time-step or the dataset as a whole.
The locations of maximum vorticity are usually associated with the most intense vortices in the dataset and often have interesting and complex
reconnection events associated with them. 
Once obtained from the service, these locations can be clustered in both 3d and 4d. This allows scientists to examine their evolution with the flow
and make subsequent analysis queries as needed in order to study these events.

Similar to the vorticity there are other fields of interest, which can be thresholded. These include ...

@@

{\bf Distributed data-parallel execution}. In most cases threshold queries operate over an entire time-step. Each such query is subdivided by 
the mediator into queries submitted to
each of the database nodes. Each node evaluates the query over the data that it has stored locally. Only a small amount of data along the boundary
need to be requested from adjacent nodes. The data are read into memory and the particular field requested is computed at each of the locations on
the grid. The same strategy applies when utilizing multiple threads per node. The norm or absolute value of the derived field at each location is compared 
against the specified threshold and if the value is higher it is maintained along with
the spatial coordinates of the location in a list. We limit the maximum size of this list to around 1,000,000 locations in order to prevent having to return
the entire time-step or a significant fraction of it for queries with thresholds that are set too low.

{\bf Application-aware cache for query results}. The central part of the evaluation strategy for threshold queries that we have developed is the 
introduction of an application-aware cache
for query results. The results of these queries are small compared to the amount of data that need to be examined and the results can be used to
answer subsequent queries as long as they are within the same region and specify the same or higher threshold. Each database node has a local 
cache. Cache entries are indexed by the field, time-step, spatial region and the threshold requested. We use a least recently
used cache replacement policy. All modifications of and queries to the cache are executed within a transaction with snapshot isolation level to
avoid dirty-reads or inconsistent view of the cache. Caching the query results
preserves the computational effort in addition to substantially reducing I/O. We do not have to derive the requested field from the raw data for 
queries that hit the cache. This results in a substantial improvement in query performance as we only have to scan
a small set of data and do not need to perform any additional computation.

@@

\begin{figure}
\centering
\includegraphics[width=3.35in]{Figures/PDF.pdf}
\caption{Probability density function of the norm of the vorticity field for a representative time-step for the MHD dataset.}
\label{fig:pdf}
\end{figure}

\subsection{Experimental Setup}
The experiments were run on the production database nodes of the JHTDB through a development Web-server hosting the Web-services. We used
the MHD dataset for the experimental runs. This dataset is partitioned across 4 database nodes according to spatial regions in the Morton z-order. 
The database nodes are 2.66 GHz dual quad-core Windows 2008 servers with SQL Server 2008 R2 and 24 GB of memory. Each node has 24 2TB 
SATA disks arranged as a set of four RAID-5 arrays. The database files are distributed in a round-robin fashion across the nodes and their associated disk
arrays. The tables storing the data are partitioned spatially along contiguous ranges of the Morton z-curve and the data for each partition reside in one
database file.

For this evaluation we looked at the performance of threshold queries to the vorticity field. The vorticity field is representative of derived fields, which
have to be computed from the stored data. The vorticity is computed from the velocity field by taking its curl:
%\begin{align}
%&\sum_{i_0=0}^{1} \sum_{i_1=0}^{1} ... \sum_{i_{d-1}=0}^{1} {(-1)}^{i_0 + i_1 + ...+ i_{d-1}}\cdot  \nonumber \\
%&\qquad s(x_0(i_0), x_1(i_1), ..., x_{d-1}(i_{d-1}))
%\label{eq:summed_volume}
%\end{align}
\begin{align}
\vec{\omega} &= \vec{\nabla} \times \vec{v} = \left( \frac{\partial}{\partial x}, \frac{\partial}{\partial y}, \frac{\partial}{\partial z} \right) 
\times \nonumber \left( v_x, v_y, v_z \right)\\
&= \left( \frac{\partial v_z}{\partial y} - \frac{\partial v_y}{\partial z}, \frac{\partial v_x}{\partial z} - \frac{\partial v_z}{\partial x}, 
\frac{\partial v_y}{\partial x} - \frac{\partial v_x}{\partial y} \right).
\label{eq:curl}
\end{align}
We use finite differencing methods of
different orders for the evaluation of the curl. For example, with 4th-order centered finite differencing each partial derivative is evaluated from the 4
adjacent grid node values as follows:  
%\begin{align}
%\frac{df}{dx}
%\label{eq:curl}
%\end{align}
\begin{align} 
\left.\frac{df}{dx}\right|_{x_n} = &\frac{2}{3\Delta x}[f(x_{n+1})-f(x_{n-1})]\nonumber\\ 
&-\frac{1}{12\Delta x}[f(x_{n+2})-f(x_{n-2})],
\end{align}
where $f$ denotes any one of the three components of the velocity and $\Delta x$ is the width of the grid in the $x$ direction.
To illustrate the distribution of the values for the norm of the vorticity field in the MHD dataset and to get an appreciation for the different threshold values
used in the experiments we show a coarse view of the probability density function for the vorticity in Fig. \ref{fig:pdf}.  
As described in section \ref{science_use_cases}
thresholding the vorticity field is important in the study of fluid dynamics and clustering the locations of maximum vorticity can lead to new insights into
the development of the most intense vortices that are observed. 
%\kk{Consider showing other fields as well; possibly thresholding the velocity by itself}

\begin{figure}
\centering
\includegraphics[width=3.35in]{Figures/cache_overhead.pdf}
\caption{Execution time for threshold queries at different threshold levels compared with the execution time of the same queries in the absence of a cache.}
\label{fig:cache_overhead}
\end{figure}

\subsection{Evaluation of cache effectiveness}
\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|c|} \hline
\multirow{2}{0.5in}{Threshold} & \multirow{2}{0.5in}{Points above threshold} & \multicolumn{3}{|c|}{Running time (s.)}\\ \cline{3-5}
& & No cache & \parbox[t]{0.55in}{With cache (miss)} & \parbox[t]{0.55in}{With cache (hit)} \\ \hline
80.0 & 4247 & 97.1 & 100.2 & 0.5\\ \hline
60.0 & 86580 & 113.7 & 115.9 & 1.2\\ \hline
44.0 & 909274 & 111.6 & 115.0 & 9.1\\ \hline
\end{tabular}
\caption{Effectiveness of caching.}
\label{tab:cache_effectiveness}
\end{table}
The central part of the strategy that we have developed for the evaluation of threshold queries of derived fields is the application-aware cache, which
stores the results of these queries. We first evaluate the overhead associated with the introduction and maintenance of the cache. Fig. \ref{fig:cache_overhead}
compares the execution time of queries in the absence of a cache with the execution time of the same queries, which interrogate the cache first.
For these experiments we requested
the locations with norms of the vorticity above thresholds at different levels. For the first set the threshold was set high (80.0) and only $\sim$4,300 points 
(or 0.0004\% of all points) were above the threshold. For the second set a medium threshold (60.0) was chosen and $\sim$87,000 points 
(or 0.0081\% of all points) were above the threshold. Finally, a low threshold (44.0) was chosen for the last set with $\sim$900,000 points 
(or 0.0847\% of all points) above the threshold. For each set a random
time-step was chosen and the queries were run against that time-step. The measurements were taken from the point of view of the end user. 
The cache was initially populated by executing several hundred unrelated queries and contained several million entries.
Cache entries for the particular time-step queried were dropped before each run, making sure that each query would produce a cache miss and would
have to be evaluated from the raw data. 
The execution times were averaged over 10 runs.
We utilized 4 threads per database
node for the evaluation of each query. As we can see from the results the overhead associated with querying the cache first is minimal, 
less than 3\% and within
the margin of error. Additionally, the method shows stable running time across different time-steps and threshold levels. The slight increase in the running
time that can be observed is due to the larger result set that has to be returned to the user.

\begin{figure}
\centering
\includegraphics[width=3.35in]{Figures/scaleup.pdf}
\caption{Execution time for threshold queries at different threshold levels -- high, medium and low. The evaluation was performed utilizing 1-8
threads per server showcasing the vertical scalability of the implementation.}
\label{fig:scaleup}
\end{figure}

@@

\subsection{Scaling and Distributed Evaluation}
The data-parallel evaluation of threshold queries exhibits good vertical scaling as shown in Fig. \ref{fig:scaleup}. For these experiments we used the same
queries and threshold values as for the runs shown in Fig. \ref{fig:cache_overhead} but with varying number of threads per node. Cache entries for the
time-step queried were again dropped before each run in order to evaluate the scaling properties of the computation of the derived field from the stored data.
The computations for all of 
the derived fields of interest (such as the vorticity) at each gird point need data from adjacent grid points only. Therefore, each node of the cluster is able to
compute the derived field from data available locally with only a little amount along the boundary of each region having to be retrieved from adjacent nodes.
Each computation is independent and can be executed in parallel. This allows us to make use of multiple threads per node.

@@

\begin{figure}
\centering
\includegraphics[width=3.35in]{Figures/io_comparison.pdf}
\caption{Execution time for threshold queries evaluated utilizing different number of
threads per server compared with the time taken to perform the I/O only.}
\label{fig:io_comparison}
\end{figure}
